Predicting Conflict Escalation

### 2023-04-29 docs/hes_final/article.qmd
Language Pass and Methods Rework

There are two principal ways in which our approach differs from previous efforts described in the literature. The first is our choice of model. Many earlier approaches, including those in O'Brien, used more traditional statistical methods, including logistic regression and actor simulations, to predict conflict. Contemporaneous work demonstrated that neural networks can outperform these traditional approaches [@10.2307/4145319]. Despite this, neural networks do not seem to have been widely adopted in conflict prediction modeling – though there are some examples, such as in Chen et al [@DBLP:conf/sac/ChenJY20]. Our work explores both approaches, and we find similar performance in some metrics, with major trade offs in other areas.

The second principal difference in our approach is our choice of input data. While newspaper and media content has been used in conflict prediction models before, it has only been included as a supplement to models that primarily rely on other features. Despite this, models that include text features have been shown to outperform equivalent models that do not [@RePEc:cam:camdae:2220]. In contrast, our approach istext-first. In other words, we made predictions primarily from text before adding other features. To achieve this, we relied on text embeddings of the news content. To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings. Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction [@RePEc:cam:camdae:2220]. These comparatively crude approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction. Our method uses text as the primary data source to make meaningful predictions. Such a model removes the need for the extensive expert surveys described in O’Brien, and overcomes the weaknesses of the strongly mean-regressing time-series models described in Mueller and Rauh.

## News media and conflict data

@@

The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/)’s team of researchers monitor thousands of sources in 20 languages to generate a dataset with monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914]. ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and  protests.

While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn. Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.

Our training data was derived from these GDELT and ACLED data. GDELT news article embeddings serve as our independent variable. Target variables for training were derived from ACLED’s conflict metrics. Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.

@@

```{python}
#| echo: false
#| label: tbl-planets
#| tbl-cap: Planets

from IPython.display import Markdown
from tabulate import tabulate
table = [["Sun",696000,1989100000],
         ["Earth",6371,5973.6],
         ["Moon",1737,73.5],
         ["Mars",3390,641.85]]
Markdown(tabulate(
  table,
  headers=["Planet","R (km)", "mass (x 10^29 kg)"]
))
```
## Defining Conflict Escalation

Our task of modeling conflict escalation required a precise definition of what conflict escalation is. Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well. [@10.1093/jeea/jvac025]

In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation. To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions. Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period.

To address this, we derived a metric designed to highlight relative “spikes” in political violence. Calculating whether a spike occurred in a given month is a two-step process. First, we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months. This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time. Next, we calculated whether that slope was in the 75th percentile of such slopes in the target country for the preceding two years. This produced a binary metric that is sensitive to small increases in political violence in peaceful countries, while also filtering out small fluctuations in long-standing conflicts.

## Methodology

In order to push beyond prior attempts at conflict prediction, our approach to modeling drew from prior work in other fields. Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events. This is unique in that it is both a Natural Language Processing (NLP) task, and a time-series task, with each vector representing both semantic meaning, and an event or observation at a point in time. This lends itself to a variety of different approaches, and we have paid particular attention to the Transformer architecture. We reviewed recent work on Transformers for time-series data, and we hypothesized that the architecture’s ability to extract long-term dependencies, as well as its ability to find complex relationships between vector embeddings made it uniquely well-suited to this task [@wen2023transformers].

We explored two approaches for getting conflict spike signals from the embeddings. In the first approach we used a fit-in-memory dataset of monthly embedding averages per country and applied statistical classification models such as K Nearest Neighbors and XGBoost. In the second approach we sampled article level embeddings and applied Transformer-based models. See the figures (X, Y) for details of both methods.

@@

The purpose of the encoder was to reduce the dimensionality of an embedding corpus in a manner analogous to a convolutional layer in a convolutional neural network. The classifier could then be used to make final predictions on a distilled dataset.

Prior to training, a base dataset was created by splitting the GDELT dataset into one $512xn$ matrix for each country and month where $n$ is the number of articles in the dataset published in that month related to that country. Each of these matrices was then split into several random subsamples of $512x50$. Samples for countries in which $n<50$ were padded. This process had two effects on training. First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample. Simultaneously, by reusing the same training label for multiple unique training samples, we were able to treat each month as though it had occurred multiple times. This greatly increased the size of the training dataset and helped overcome the limitations imposed by the limited time window available in the GDELT embedding corpus.

This new dataset represents a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive less. Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons.

With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512x50$ matrices, this time without resampling. Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector. Each vector was then stacked, resulting in a single matrix for each country and month of size $16x\frac{n}{50}$. This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target.

@@

{{< embed figs.ipynb#tbl-best-max-metrics >}}

{{< embed figs.ipynb#tbl-best-min-metrics >}}

@@

{{< embed figs.ipynb#fig-roc >}}

{{< embed figs.ipynb#fig-prc >}}

@@

In this work we explored the possibility of predicting future geo-political conflict escalation at a country level
using a global collection of news publications. We explored different ways of aggregating the data and
applied both statistical machine learning and deep learning transformer architecture models.
The results suggest that text embeddings provide a strong enough signal to make such predictions.
Different data sampling and modeling approaches produce comparable results based on common metrics. 
Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.

@@

We think there are many ways to extend this work. We can group them to align with our initial research and
practical objectives.

@@

On the transormers side, the output of the encoder also represented a reduced dimension dataset
further fed into the second transformer, the classifier. Using the output of the encoder as input to other
classifiers could be another direction of research.

@@

Building a more robust data processing pipeline is another important direction. This would involve
designing either a batch or streaming system that can reliably collect data, preprocess
and aggregate it into required format, train the model and make predictions. The system should handle 
computational resources and cost constraints.

>>>>

## Methodology
The principal difference between our approach and that of prior work is in our choice of predictive features, which we extract directly from text embeddings of publicly available news coverage. While newspaper and media content has been used in conflict prediction models before, it is generally only included as a supplement to models that primarily rely on other features. Despite this, models that include text features have been shown to outperform equivalent models that do not [@RePEc:cam:camdae:2220]. 
In contrast, our approach is text-first, making predictions primarily or exclusively from text, with few or no other features added. To achieve this, we relied on text embeddings of the news content. To the best of our knowledge, no conflict prediction model described in the literature makes use of text embeddings. Previous efforts have included statistical analysis of text, and unsupervised Latent Dirichlet Allocation (LDA) for topic extraction [@RePEc:cam:camdae:2220]. These approaches to Natural Language Processing (NLP) have relegated text data to a peripheral role in conflict prediction. Our method uses text as the primary data source to make meaningful predictions. Such a model removes the need for the extensive expert surveys described in O’Brien, and overcomes the weaknesses of the strongly mean-regressing time-series models described in Mueller and Rauh.

In order to elevate text data to the center-stage of our prediction efforts, we drew from prior work in other fields. Our training data was structured such that each document is represented as an embedding vector, and multiple documents are combined with a label to create an aggregate picture of recent events. 
We explored two approaches for extracting conflict signals from text embeddings. In the first approach we used a fit-in-memory dataset of monthly embedding averages per country per month and applied statistical classification models such as K Nearest Neighbors and XGBoost. In the second approach we stacked multiple article-level embeddings and applied Transformer-based models to extract a signal directly from the text corpus. See the figures (X, Y) for details of both methods.

### Defining Conflict Escalation
Our task of modeling conflict escalation required a precise definition of what conflict escalation is. We began with raw event and casualty counts from The [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/) , whose team of researchers monitor thousands of sources in 20 languages to generate a dataset containing monthly totals of conflict events and their related fatalities [@doi:10.1177/0022343310378914]. ACLED defines these events as incidents involving armed groups with political objectives as well as demonstrations, riots, and protests.
With these totals in-hand, we needed to create a target to label our training data. Prior work has treated this as a binary classification task, asking whether or not conflict escalated in a given month, and we adopt this approach as well. [@10.1093/jeea/jvac025]
In contrast to prior work, however, we do not treat the simple presence of a fatality due to conflict as sufficient to indicate conflict escalation. To achieve our objective of creating a generalizable conflict model, we required a metric that was consistent across varying local conditions. Small changes in the absolute number of monthly fatalities can have widely different implications depending on the country and time period. 
To address this, we derived a metric designed to highlight relative “spikes” in political violence. Calculating whether a spike occurred in a given month is a two-step process. First, we calculated the trend in violence for a given month, defined by the slope of an ordinary least squares regression of the number of ACLED-documented fatalities that occurred in the target month, and the preceding two months. This trend allowed us to estimate whether violence was increasing or decreasing in the target country at any given time. Next, we calculated whether that slope was in the $75^{th}$ percentile of such slopes in the target country for the preceding two years. This produced a binary metric that is sensitive to small increases in political violence in peaceful countries, while also filtering out small fluctuations in long-standing conflicts.

### News Content Data

@@

While many of GDELT’s databases reflect news articles from 1 January 1979 onward, GDELT’s news article embeddings exist only from 1 January 2020, limiting the time period for which training data can be drawn. Availability of ACLED’s monthly event and fatality totals varies by country, but event and fatality totals are available for all countries since at least 1 January 2020.
Our training data was derived from these GDELT and ACLED data. GDELT news article embeddings served as our independent variable, while the “spike” metric derived from ACLED’s conflict data served as our target variable. Hard dates were set for train, test, and validation splits, and none of our models saw data published after October 2022 during training, with the remaining months used for testing and validation.

@@

@@

The purpose of the encoder was to reduce the dimensionality of the embedding corpus by analyzing a subset of article embeddings and returning a single vector of substantially reduced size. This is roughly analogous to the role of a convolutional layer in a CNN. The classifier could then be used to make final predictions on the resulting distilled dataset.

To prepare the training dataset, our full sample of the GDELT corpus was split into a single $512xn$ matrix for each country and month where $n$ is the number of articles published in that month related to that country. Each of these matrices was then split into several random subsamples of 50 articles each. Samples for countries in which $n<50$ were padded to $512x50$. This process had two effects on training. First, by limiting the size of each sample to 50 articles, we reduced the dimensionality of each sample. Simultaneously, by splitting the corpus into multiple unique training examples per country per month, we were able to substantially increase the size of the training dataset.

This new dataset offered major advantages in combatting the curse of dimensionality, however, it also represented a distorted view of the world, which overrepresents countries that receive more media coverage and underrepresents those that receive les, amplifying the biases already present in the GDELT dataset. Despite this limitation, this dataset was sufficient to train the encoder, which was fit to the modified dataset with the time-shifted spike metric as its target. Once training was complete, the final layer of the encoder was removed, exposing the penultimate layer of 16 dense neurons. The output of these 16 neurons would be used in the classification stage.

With the encoder's training complete, we returned to the original base dataset. Each country-month matrix was again split into $512x50$ matrices. Each of these samples could then be passed into the encoder, which distilled it into a $length-16$ vector. Each vector was then stacked, resulting in a single training example for each country and month of size $16x\frac{n}{50}$. This produced a new, distilled, and undistorted dataset which could then be used to train the classifier using the same spike metric as its target. 
In contrast to our statistical modeling approach, the transformer models are trained purely on text embeddings. The model was not conditioned by country, and no additional features were added beyond the text corpus itself.

@@

@@

@@

In this work we explored the possibility of predicting future geo-political conflict escalation at a country level
using a global collection of news publications. We explored different ways of aggregating the data and
applied both statistical machine learning and deep learning models.
The results suggest that text embeddings provide a strong enough signal to make such predictions.
Different data sampling and modeling approaches produce comparable results based on common metrics. 
Transformer-based model generalizes better across geographies while statistical model provides a size/cost of deployment advantage.

@@

We foresee several ways to extend this work, and there are exciting opportunities for both further research and operationalizations.

@@

The transformer models have a great deal of room for hyperparameter tuning. Further experimentation with the size of the encoder output, the number of articles per training sample, and implementing overlapping strides during distillation could all improve the downstream performance of the classifier. It may also be possible to use the output of the encoder as input to some other classification model architecture.

@@

Building a more robust and automated data processing pipeline is another important direction. This would involve
designing either a batch or streaming system that can reliably collect data, preprocess
and aggregate it into required format, train the model and make predictions. The system should handle 
computational resources and cost constraints.
